Research Notes

### 2018-11-12 writeups/notes.tex
writeup simplified anelastic

\section{Traveling Breaking Front}

@@

@@

\section{The Anelastic/Boussinesq Approximations}

@@

>>>>

@@

\chapter{Anelastic Fluid Equations}

\section{Inadequacy of Incompressible Stratified Equations}

To perform simulations up to $u \sim c_p \ll c_s$ and include the advective
terms but no sound-speed-order terms, we need to make sure our equations
qualitatively match the fully nonlinear (including sound speed) equations.
In a stratified atmosphere, the incompressible equations are qualitatively
wrong. The full fluid equations consist of conservation equations plus an
equation of state, so our reduced equations must conserve the same quantities.
The equations which we simulated above, the incompressible nonlinear equations,
must themselves also conserve the same quantities.

Our system, which drops its energy conservation equation, consists of momentum
conservation, mass conservation and an incompressible equation of state
$\rd{\rho}{t} = 0$. Note however that the horizontally averaged horizontal
momentum $\ev*{\rho u_{1x}} = \ev*{\rho_0 u_{1x}}$ is \emph{not} conserved by
our above equations:
\begin{equation}
    \rho_0 \pd{u_{1x}}{t} + \rho_0 \p*{\vec{u}_1 \cdot \vec{\nabla}} u_{1x}
        + \vec{\nabla}P_1 = 0.
\end{equation}
Since $\pd{\rho_0}{t} = 0$, we may try to put this into flux conservative form
via
\begin{equation}
    \pd{(\rho_0 u_{1x})}{t} + \vec{\nabla} \cdot
        \p*{\rho_0 \vec{u}_1 u_{1x} + \bm{P}_1}
        = u_{1x} \vec{\nabla} \cdot \p*{\rho_0 \vec{u}_1}.
\end{equation}
We combine some scalar gradients into tensors, but it's clear that there is an
extra term that \emph{is not} cancelled by the usual incompressibility criterion
$\vec{\nabla} \cdot \vec{u}_1 = 0$. It bears noting this term only cancels
\emph{in the Boussinesq approximation} where $\rho_0$ is held constant.

\section{Anelastic Fluid Equations}

The anelastic fluid equations replace incompressibility with the obvious
$\vec{\nabla} \cdot (\rho_0\vec{u}_1) = 0$ to conserve momentum $\rho_0 u_{1x}$
exactly. There are however many flavors of anelastic approximations however, the
large majority of which reintroduce a temperature/entropy variable. This
probably comes about because most uses of the anelastic equations examine
convective systems, for which thermal diffusivity is important. For us, where
all fluid motion is fairly local, it seems reasonable to only permit fluid
advection and simplify our model, so we can stay with only two state variables
$P, \rho$ along with the flow $\vec{u}$. We still use numerical viscosities, but
these are more so a subgrid model and not physical, so we will just reintroduce
them before simulating.

From here, it's a simple matter of taking limits of existing anelastic
approximations. We will start with the full system from Glatzmaier's book:
\begin{align}
    0 &= \vec{\nabla} \cdot \p*{\rho_0 \vec{u}_1},\\
    \rho_1 &= \frac{\rho_0}{\gamma P_0}P_1 - \frac{\rho_0}{c_p}S_1
        = \at{\pd{\rho_0}{P_0}}_S P_1 + \at{\pd{\rho_0}{S_0}}_P S_1,\\
    N^2 &= \frac{g}{c_p}\rd{S_0}{z},\\
    0 &= \pd{\vec{u}_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
        + \vec{\nabla}\p*{\frac{P_1}{\rho_0} + \Phi_1}
        + \at{\pd{\rho_0}{S_0}}P \frac{g}{\rho_0}S\hat{z}
        - \nu \p*{\nabla^2 \vec{u}_1 + \frac{1}{3}\vec{\nabla}\p*{\vec{\nabla}
            \cdot \vec{u}_1}},\\
    0 &= \pd{S_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}S_1
        + u_{1z}\rd{S_0}{z}
        - \frac{1}{\rho_0 T_0} \vec{\nabla} \cdot \p*{\kappa \rho_0 T_0
            \vec{\nabla} S_1}.
\end{align}
We obviously take $\Phi_1 = 0$ the Cowling approximation. Of course, in our
problem, we may drop all the viscous terms from the momentum equation and energy
equation, which immediately leaves us with
\begin{align}
    \vec{\nabla} \cdot \p*{\rho_0 \vec{u}_1} &= 0,\\
    \pd{\vec{u}_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
        + \vec{\nabla} \p*{\frac{P_1}{\rho_0}} - \frac{g}{c_p}S_1 \hat{z}
            &= 0,\\
    \pd{S_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}S_1
        &= 0.
\end{align}
Note that when accounting for $N^2 = \frac{g}{c_p}\rd{S_0}{z}$ so $c_p = S_0$
immediately gives us Sutherland's equations from his 2011 paper.

We now make one further approximation: we have already taken the low Mach number
approximation, so we will approximate $\rho_1 = \frac{\rho_0}{\gamma P_0}P_1 -
\frac{\rho_0}{S_0}S_1 \approx -\frac{\rho_0}{S_0}S_1$, which allows us to
eliminate $S$. We must be careful in taking this approximation (and indeed, it
may be unjustified), since the difference term $\vec{\nabla}\frac{P_1}{\rho_0} -
\frac{\vec{\nabla}P_1}{\rho_0} = -\frac{P_1 \vec{\nabla}\rho_0}{\rho_0^2}$ is
now small as well. What to do with it is dictated by the requirement that
horizontal momentum be conserved. With this in mind, the momentum equation
must be
\begin{equation}
    \pd{\vec{u}_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
        + \frac{\vec{\nabla}P_1}{\rho_0} + \frac{\rho_1}{\rho_0}g\hat{z} = 0.
\end{equation}
We may verify that horizontal momentum is conserved by multiplying through with
$\rho_0$ again.

The entropy equation takes a bit more work, but we may show that
\begin{align}
    -\frac{c_p}{\rho_0}\pd{\rho_1}{t} - c_p \vec{u}_1 \cdot
        \vec{\nabla}\p*{\frac{\rho_1}{\rho_0}} - \frac{u_1 c_p N^2}{g} &= 0,\\
    \pd{\rho_1}{t} + \vec{u}_1 \cdot \vec{\nabla}\rho_1
        - \frac{\rho_1\vec{u}_1 \cdot \vec{\nabla}\rho_0}{\rho_0}
        - \frac{u_{1z}\rho_0}{H} &= 0,\nonumber\\
    \pd{\rho_1}{t} + \vec{\nabla}\p*{\rho_1 \vec{u}_1} - \frac{u_{1z}\rho_0}{H}
        &= 0.
\end{align}
We have used $\vec{\nabla} \cdot \p*{\rho_0 \vec{u}_1} = 0$.

@@

\section{The Anelastic/Boussinesq Approximations (old)}

@@

This seems to be largely dated now, we believe that the sharp jump in the
nonlinear simulations we are seeing are mean flow interactions, rather than
hydraulic jumps.

\subsection{Traveling Breaking Front}
